CommonMind Consortium: DLPFC

### 2021-01-14 combined_analysis/combine_all_DLPFC.Rmd
For public release

library(githubr)
parentId = 'syn22416298';
activityName = 'Joint analysis of DLPFC count data';
activityDescription = 'Joint analysis of DLPFC count data';

@@

library(ggplot2)
library(dplyr)
library(ComplexHeatmap)
library(readxl)
library(gtools)
library(corrplot)
library(fdrtool)
library(pvclust)
library(RColorBrewer)
library(metafor)
library(data.table)
library(cowplot)
library(ggraph)
library(tidygraph)
library(kableExtra)
library(githubr)
library(ggrepel)
library(corpcor)
library(corrplot)
library(zenith)
library(compositions)
library(knitr)
library(patchwork)
library(plyr) 
library(BiocParallel)

@@

```{r read.counts}
# Read raw counts from MPP and HBCC
counts.MPP = fread(synGet('syn21867938')$path, data.table=FALSE)
rownames(counts.MPP) = counts.MPP$Geneid
counts.MPP = counts.MPP[,-c(1:6)]

# HBCC - DLFPC
counts.HBCC = fread(synGet('syn21886235')$path, data.table=FALSE)
rownames(counts.HBCC) = counts.HBCC$Geneid
counts.HBCC = counts.HBCC[,-c(1:6)]

# # HBCC - sgACC
counts.HBCC4 = fread(synGet('syn21867866')$path, data.table=FALSE)
rownames(counts.HBCC4) = counts.HBCC4$Geneid
counts.HBCC4 = counts.HBCC4[,-c(1:6)]

geneCountsMerged = cbind(counts.MPP, counts.HBCC, counts.HBCC4)
```

### Data download
#### Obtain effective count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
downloadFile <- function(id){
  fread(synGet(id)$path, data.table = F)
}
downloadFile_version <- function(id , version){
  fread(synGet(id, version = version)$path, data.table = F)
}

# Get RNASeq QCmetadata
METADATA_QC_DLPFC_ID = 'syn16816488' 
ALL_USED_IDs = METADATA_QC_DLPFC_ID
metadata = downloadFile_version(METADATA_QC_DLPFC_ID, version = 19) %>%
  dplyr::select(`Individual_ID`, Institution, Cohort, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 'pH',
                Sample_RNA_ID, one_of('rnaSeq_isolation:RIN',
                                      'rnaSeq_dissection:Brain_Region', 
                                      'rnaSeq_isolation:RNA_Isolation_Batch',
                                      'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
                                      'rnaSeq_report:Flowcell_Batch', 'rnaSeq_report:Exclude?',
                                      'rnaSeq_report:Mapped_Reads',
                                      'rnaSeq_report:Intragenic_Rate','rnaSeq_report:Intronic_Rate',
                                      'rnaSeq_report:Intergenic_Rate','rnaSeq_report:Genes_Detected',
                                      'rnaSeq_report:Expression_Profiling_Efficiency',
                                      'rnaSeq_report:rRNA_Rate',
                                      'rnaSeq_report:Total_Reads', 'rnaSeq_report:Percent_Aligned','rnaSeq_report:Transcripts_Detected'))

# Get ancestry vector calculated using gemtools
ANCESTRY_ID = 'syn9922992'
ALL_USED_IDs = c(ALL_USED_IDs, ANCESTRY_ID)
ANCESTRY.HBCC = downloadFile(ANCESTRY_ID) %>% 
  plyr::rename(c(ID = 'SNP_report:Genotyping_Sample_ID'))

ANCESTRY_ID = 'syn2511399'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = ANCESTRY_ID
ANCESTRY.MPP = downloadFile(ANCESTRY_ID) %>% 
  plyr::rename(c('DNA_report..Genotyping.Sample_ID' = 'SNP_report:Genotyping_Sample_ID'))

ANCESTRY = rbind(ANCESTRY.HBCC[,colnames(ANCESTRY.MPP)[-2]], ANCESTRY.MPP[,colnames(ANCESTRY.MPP)[-2]])

# Get genotype ids from synapse. 
GENOTYPE_ID = 'syn16816490'
ALL_USED_IDs = c(ALL_USED_IDs, GENOTYPE_ID)
GENOTYPE = downloadFile(GENOTYPE_ID) %>% 
  dplyr::select(Individual_ID, `SNP_report:Genotyping_Sample_ID`, `SNP_report:Exclude?`) %>% 
  dplyr::inner_join(ANCESTRY) %>% 
  dplyr::filter(is.na(`SNP_report:Exclude?`))
```

### Data preprocessing
```{r preprocess.data}
# Merge metadata 
METADATA = metadata %>%
  dplyr::left_join(GENOTYPE) %>%      
  dplyr::rename(Region = `rnaSeq_dissection:Brain_Region`,
                PMI = `PMI_(in_hours)`,
                RIN = `rnaSeq_isolation:RIN`, 
                ReportExclude = `rnaSeq_report:Exclude?`,
                GenotypeExclude = `SNP_report:Exclude?`,
                SampleID = Sample_RNA_ID,
                LibraryBatch = `rnaSeq_report:Library_Batch`,
                FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
                IsolationBatch = `rnaSeq_isolation:RNA_Isolation_Batch`, #isolation batch to replace ribozero batch
                RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
                MappedReads = `rnaSeq_report:Mapped_Reads`,
                IntragenicRate = `rnaSeq_report:Intragenic_Rate`, IntronicRate = `rnaSeq_report:Intronic_Rate`, 
                IntergenicRate = `rnaSeq_report:Intergenic_Rate`, GenesDetected = `rnaSeq_report:Genes_Detected`,
                ExpProfEfficiency = `rnaSeq_report:Expression_Profiling_Efficiency`, rRNARate = `rnaSeq_report:rRNA_Rate`,
                TotalReads = `rnaSeq_report:Total_Reads`, AlignmentRate = `rnaSeq_report:Percent_Aligned`, TranscriptsDetected = `rnaSeq_report:Transcripts_Detected`) %>%
  dplyr::select(SampleID, Individual_ID, Institution, Cohort, Reported_Gender,Sex, Ethnicity, ageOfDeath, PMI, Dx, Region,
                RIN, ReportExclude, GenotypeExclude, EV.1, EV.2, EV.3, EV.4, EV.5, pH,
                LibraryBatch, FlowcellBatch, IsolationBatch, RibozeroBatch, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate, TranscriptsDetected) %>%
  		 		 dplyr::filter(SampleID %in% colnames(geneCountsMerged), !is.na(SampleID)) %>%
  		 		 dplyr::filter(Region %in% c('DLPFC')) %>%
			 	 dplyr::mutate(Cohort = forcats::fct_recode(Cohort, `MSSM-Penn-Pitt` = 'MSSM-Penn-Pitt', `NIMH-HBCC`='NIMH-HBCC', `NIMH-HBCC`='NIMH-HBCC-4')) %>%
			 	 dplyr::filter(Dx %in% c('Control', "SCZ"))  %>%
         dplyr::filter(ageOfDeath > 17)

@@

## t-statistics for CASKIN1
```{r CASKIN1}
 df_two['ENSG00000167971',]  %>% kable(caption="t-statistics for CASKIN1", row.names=FALSE) %>% kable_styling(full_width=FALSE)
```

@@

```{r store, eval=TRUE}
if( params$upload ){    
  # Code
  # Make sure these match the folder name and parent Id in the first chunk.
  CODE <- Folder(name = "Joint analysis of DLPFC count data", parentId = "syn22416298")
  CODE <- synStore(CODE)

@@

  # Sex by Disease interaction
  storeData( "fit_inter_sex.disease.lst", "dream fit for interaction analysis")
  storeData( "res_zenith.inter_sex.disease", "zenith results for interaction analysis")
  storeData( "res_meta_zenith.inter_sex.disease", "Meta-analysis of zenith results for interaction analysis")
  storeData( "res_meta_zenith.inter_sex.disease_autosome", "Meta-analysis of zenith results for interaction analysis -  autosomal genes only")
  storeData( "df_DE_results.inter_sex.disease", "Differential expression result for each cohort for interaction analysis")
  storeData( "df_meta.inter_sex.disease", "Meta-analysis for differential expression for interaction analysis")
}

#   rmarkdown::render("combined_analysis/combine_all_DLPFC.Rmd")

>>>>

library(githubr)
parentId = 'syn24172641';
activityName = 'Joint analysis of DLPFC count data';
activityDescription = 'Joint analysis of DLPFC count data';

@@

library(ggplot2)
library(dplyr)
library(ComplexHeatmap)
library(readxl)
library(gtools)
library(corrplot)
library(fdrtool)
library(ashr)
library(pvclust)
library(RColorBrewer)
library(metafor)
library(data.table)
library(cowplot)
library(ggraph)
library(tidygraph)
library(kableExtra)
library(githubr)
library(ggrepel)
library(corpcor)
library(corrplot)
library(zenith)
library(compositions)
library(knitr)
library(patchwork)
library(plyr) 
library(BiocParallel)

@@

```{r read.counts}
# Read raw counts from MPP and HBCC
# counts.MPP = fread(synGet('syn21867938')$path, data.table=FALSE)

ALL_USED_IDs = c()

# Just DLPFC counts
ALL_USED_IDs = c(ALL_USED_IDs, 'syn24172729')
counts.MPP = fread(synGet('syn24172729')$path, data.table=FALSE)
rownames(counts.MPP) = counts.MPP$Geneid
counts.MPP = counts.MPP[,-c(1:6)]

# HBCC - DLFPC
ALL_USED_IDs = c(ALL_USED_IDs, 'syn21886235')
counts.HBCC = fread(synGet('syn21886235')$path, data.table=FALSE)
rownames(counts.HBCC) = counts.HBCC$Geneid
counts.HBCC = counts.HBCC[,-c(1:6)]

# # HBCC - sgACC
# NOT INCLUDED in release
# counts.HBCC4 = fread(synGet('syn21867866')$path, data.table=FALSE)
# rownames(counts.HBCC4) = counts.HBCC4$Geneid
# counts.HBCC4 = counts.HBCC4[,-c(1:6)]

# geneCountsMerged = cbind(counts.MPP, counts.HBCC, counts.HBCC4)
geneCountsMerged = cbind(counts.MPP, counts.HBCC)
```

### Data download
#### Obtain effective count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
downloadFile <- function(id, version=NULL){
  fread(synGet(id, version = version)$path, data.table = F)
}
```

```{r get.metadata, eval=FALSE}
# Get RNASeq QCmetadata
# METADATA_QC_DLPFC_ID = 'syn16816488'; version = 19  # private version
# # METADATA_QC_DLPFC_ID = 'syn18358379'; version = NULL  # public version
# ALL_USED_IDs = METADATA_QC_DLPFC_ID
# metadata = downloadFile_version(METADATA_QC_DLPFC_ID, version = version) %>%
#   dplyr::select(`Individual_ID`, Institution, Cohort, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 'pH',
#                 Sample_RNA_ID, one_of('rnaSeq_isolation:RIN',
#                                       'rnaSeq_dissection:Brain_Region', 
#                                       'rnaSeq_isolation:RNA_Isolation_Batch',
#                                       'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
#                                       'rnaSeq_report:Flowcell_Batch', 'rnaSeq_report:Exclude?',
#                                       'rnaSeq_report:Mapped_Reads',
#                                       'rnaSeq_report:Intragenic_Rate','rnaSeq_report:Intronic_Rate',
#                                       'rnaSeq_report:Intergenic_Rate','rnaSeq_report:Genes_Detected',
#                                       'rnaSeq_report:Expression_Profiling_Efficiency',
#                                       'rnaSeq_report:rRNA_Rate',
#                                       'rnaSeq_report:Total_Reads', 'rnaSeq_report:Percent_Aligned','rnaSeq_report:Transcripts_Detected'))

# # Get ancestry vector calculated using gemtools
# ANCESTRY_ID = 'syn9922992'
# ALL_USED_IDs = c(ALL_USED_IDs, ANCESTRY_ID)
# ANCESTRY.HBCC = downloadFile(ANCESTRY_ID) %>% 
#   plyr::rename(c(ID = 'SNP_report:Genotyping_Sample_ID'))

# ANCESTRY_ID = 'syn2511399'
# ALL_USED_IDs[length(ALL_USED_IDs)+1] = ANCESTRY_ID
# ANCESTRY.MPP = downloadFile(ANCESTRY_ID) %>% 
#   plyr::rename(c('DNA_report..Genotyping.Sample_ID' = 'SNP_report:Genotyping_Sample_ID'))

# ANCESTRY = rbind(ANCESTRY.HBCC[,colnames(ANCESTRY.MPP)[-2]], ANCESTRY.MPP[,colnames(ANCESTRY.MPP)[-2]])

# # Get genotype ids from synapse. 
# GENOTYPE_ID = 'syn16816490'
# ALL_USED_IDs = c(ALL_USED_IDs, GENOTYPE_ID)
# GENOTYPE = downloadFile(GENOTYPE_ID) %>% 
#   dplyr::select(Individual_ID, `SNP_report:Genotyping_Sample_ID`, `SNP_report:Exclude?`) %>% 
#   dplyr::inner_join(ANCESTRY) %>% 
#   dplyr::filter(is.na(`SNP_report:Exclude?`))

# # Merge metadata   
# METADATA = metadata %>%
#   dplyr::left_join(GENOTYPE) %>%      
#   dplyr::rename(Region = `rnaSeq_dissection:Brain_Region`,
#                 PMI = `PMI_(in_hours)`,
#                 RIN = `rnaSeq_isolation:RIN`, 
#                 ReportExclude = `rnaSeq_report:Exclude?`,
#                 GenotypeExclude = `SNP_report:Exclude?`,
#                 SampleID = Sample_RNA_ID,
#                 LibraryBatch = `rnaSeq_report:Library_Batch`,
#                 FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
#                 IsolationBatch = `rnaSeq_isolation:RNA_Isolation_Batch`, #isolation batch to replace ribozero batch
#                 RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
#                 MappedReads = `rnaSeq_report:Mapped_Reads`,
#                 IntragenicRate = `rnaSeq_report:Intragenic_Rate`, IntronicRate = `rnaSeq_report:Intronic_Rate`, 
#                 IntergenicRate = `rnaSeq_report:Intergenic_Rate`, GenesDetected = `rnaSeq_report:Genes_Detected`,
#                 ExpProfEfficiency = `rnaSeq_report:Expression_Profiling_Efficiency`, rRNARate = `rnaSeq_report:rRNA_Rate`,
#                 TotalReads = `rnaSeq_report:Total_Reads`, AlignmentRate = `rnaSeq_report:Percent_Aligned`, TranscriptsDetected = `rnaSeq_report:Transcripts_Detected`) %>%
#   dplyr::select(SampleID, Individual_ID, Institution, Cohort, Reported_Gender,Sex, Ethnicity, ageOfDeath, PMI, Dx, Region,
#                 RIN, ReportExclude, GenotypeExclude, EV.1, EV.2, EV.3, EV.4, EV.5, pH,
#                 LibraryBatch, FlowcellBatch, IsolationBatch, RibozeroBatch, MappedReads, TotalReads, GenesDetected, AlignmentRate,
#                 IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate, TranscriptsDetected) 
```

```{r get.metadata.2}
# Get ancestry vector calculated using gemtools
ANCESTRY_ID = 'syn17894713'
ALL_USED_IDs = c(ALL_USED_IDs, ANCESTRY_ID)
ANCESTRY.HBCC = downloadFile(ANCESTRY_ID) %>% 
  dplyr::rename(ID = 'Genotyping_Sample_ID')
ANCESTRY_ID = 'syn17346100'
ALL_USED_IDs = c(ALL_USED_IDs, ANCESTRY_ID)
ANCESTRY.MPP = downloadFile(ANCESTRY_ID) %>% 
  dplyr::rename('DNA_report..Genotyping.Sample_ID' = 'Genotyping_Sample_ID')
ANCESTRY = rbind(
  ANCESTRY.HBCC[,colnames(ANCESTRY.MPP)[-2]], 
  ANCESTRY.MPP[,colnames(ANCESTRY.MPP)[-2]]
  )
# Get genotype ids from synapse. 
GENOTYPE_ID = "syn18358480"
ALL_USED_IDs = c(ALL_USED_IDs, GENOTYPE_ID)
GENOTYPE = downloadFile(GENOTYPE_ID) %>% 
  dplyr::select(Individual_ID, `Genotyping_Sample_ID`, `Exclude`) %>% 
  dplyr::inner_join(ANCESTRY) %>% 
  dplyr::filter(is.na(`Exclude`))
# Get RNASeq QCmetadata
CLINICAL_ID = "syn3354385"
clinical = downloadFile(CLINICAL_ID)
ASSAY_ID = "syn24173489"
rnaseq = downloadFile(ASSAY_ID)
metadata <- right_join(
  clinical, 
  rnaseq,
  )
ALL_USED_IDs = c(ALL_USED_IDs,CLINICAL_ID, ASSAY_ID)
metadata = metadata %>%
  dplyr::select(
    `Individual ID`, Institution, Cohort, `Reported Gender`, Sex, Ethnicity, 
    `Age of Death`, `PMI (in hours)`, Dx, pH, SampleID, one_of(
      'RIN','Brain_Region', 'Ribozero_Batch', 'Library_Batch', 
      'Flowcell_Batch', 'Mapped_Reads', 'Intragenic_Rate','Intronic_Rate', 
      'Intergenic_Rate','Genes_Detected', 'Expression_Profiling_Efficiency', 
      'rRNA_Rate', 'Total_Reads', 'Percent_Aligned', 'Transcripts_Detected',
      'Exclude?')
    ) %>% 
  dplyr::filter(`Brain_Region` %in% c("DLPFC"))
metadata$`Age of Death`[metadata$`Age of Death`=="90+"] = 90
metadata$`Age of Death` = as.numeric(metadata$`Age of Death`)

# Merge metadata 
METADATA = metadata %>%
  dplyr::left_join(GENOTYPE, by = c("Individual ID" = "Individual_ID")) %>%  
  dplyr::rename(
    Region = `Brain_Region`,
    PMI = `PMI (in hours)`,
    RIN = `RIN`,
    ReportExclude = `Exclude?`,
    GenotypeExclude = Exclude,
    LibraryBatch = `Library_Batch`,
    FlowcellBatch = `Flowcell_Batch`,
    RibozeroBatch = `Ribozero_Batch`,
    MappedReads = `Mapped_Reads`,
    IntragenicRate = `Intragenic_Rate`, 
    IntronicRate = `Intronic_Rate`, 
    IntergenicRate = `Intergenic_Rate`, 
    GenesDetected = `Genes_Detected`,
    ExpProfEfficiency = `Expression_Profiling_Efficiency`, 
    rRNARate = `rRNA_Rate`,
    TotalReads = `Total_Reads`, 
    AlignmentRate = `Percent_Aligned`, 
    TranscriptsDetected = `Transcripts_Detected`,    
    TranscriptsDetected = `Transcripts_Detected`,    
    Reported_Gender = `Reported Gender`,
    ageOfDeath = `Age of Death`,
    IndividualID = `Individual ID`) %>%
  dplyr::select(
    SampleID, IndividualID, Institution, Cohort,Reported_Gender, Sex, 
    Ethnicity, ageOfDeath, PMI, Dx, RIN, EV.1, EV.2, EV.3, EV.4, EV.5, 
    LibraryBatch, FlowcellBatch, RibozeroBatch, MappedReads, TotalReads, 
    GenesDetected, AlignmentRate, IntragenicRate, IntergenicRate, IntronicRate, 
    ExpProfEfficiency, rRNARate, TranscriptsDetected, ReportExclude, 
    GenotypeExclude, pH)
```

### Data preprocessing
```{r preprocess.data}

METADATA = METADATA %>%
  		 		 dplyr::filter(SampleID %in% colnames(geneCountsMerged), !is.na(SampleID)) %>%
  		 		 # dplyr::filter(Region %in% c('DLPFC')) %>%
			 	 dplyr::mutate(Cohort = forcats::fct_recode(Cohort, `MSSM-Penn-Pitt` = 'MSSM-Penn-Pitt', `NIMH-HBCC`='NIMH-HBCC')) %>% # , `NIMH-HBCC`='NIMH-HBCC-4'
			 	 dplyr::filter(Dx %in% c('Control', "SCZ"))  %>%
         dplyr::filter(ageOfDeath > 17)

@@

## t-statistics for CASKIN1
```{r CASKIN1}
 df_two['ENSG00000167971',]  %>% kable(caption="t-statistics for CASKIN1", row.names=FALSE) %>% kable_styling(full_width=FALSE)
 df_two['ENSG00000163634',]  %>% kable(caption="t-statistics for CASKIN1", row.names=FALSE) %>% kable_styling(full_width=FALSE)
```

@@

```{r store, eval=FALSE}
if( params$upload ){    
  # Code
  # Make sure these match the folder name and parent Id in the first chunk.
  CODE <- Folder(name = "Joint analysis of DLPFC count data", parentId = "syn24172641")
  CODE <- synStore(CODE)

@@

  # Sex by Disease interaction
  storeData( "fit_inter_sex.disease.lst", "dream fit for interaction analysis")
  storeData( "res_zenith.inter_sex.disease", "zenith results for interaction analysis")
  storeData( "res_meta_zenith.inter_sex.disease", "Meta-analysis of zenith results for interaction analysis")
  storeData( "res_meta_zenith.inter_sex.disease_autosome", "Meta-analysis of zenith results for interaction analysis -  autosomal genes only")
  storeData( "df_DE_results.inter_sex.disease", "Differential expression result for each cohort for interaction analysis")
  storeData( "df_meta.inter_sex.disease", "Meta-analysis for differential expression for interaction analysis")
}
```
